---
title: "Mapping Sequencing Data to Genome Annotation and Visualization"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi = 120
)

pkg_ok <- function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
}

need <- function(pkgs) {
  all(vapply(pkgs, pkg_ok, logical(1)))
}

missing_pkgs <- function(pkgs) {
  pkgs[!vapply(pkgs, pkg_ok, logical(1))]
}

erbs_peaks <- NULL
annotated_peaks <- NULL
gm12878_peaks <- NULL
txdb <- NULL
all_genes <- NULL
all_exons <- NULL
promoter_ranges <- NULL
tss_ranges <- NULL
esrra_neigh <- NULL
annotated_esrra_window_peaks <- NULL
esrra_window_peaks <- NULL
```

# Tutorial

This is a practical ERBS tutorial. It starts with the ERBS peak file, turns it into a GRanges object, annotates every peak against the human `hg19` genome, and then visualizes the ESRRA locus with Gviz.

This notebook is written to be runnable when the relevant Bioconductor packages are installed.

```{r}
core_packages <- c(
  "knitr",
  "BiocManager",
  "GenomicRanges",
  "GenomeInfoDb",
  "IRanges",
  "S4Vectors",
  "rtracklayer",
  "ERBS",
  "BSgenome",
  "BSgenome.Hsapiens.UCSC.hg19",
  "TxDb.Hsapiens.UCSC.hg19.knownGene",
  "Homo.sapiens",
  "AnnotationDbi",
  "GenomicFeatures",
  "Biostrings",
  "Gviz",
  "ggplot2"
)
```

```{r}
package_status <- data.frame(
  package = core_packages,
  installed = vapply(core_packages, pkg_ok, logical(1)),
  row.names = NULL
)

knitr::kable(package_status)
```

TRUE means the package is available. Any FALSE row marks a package that should be installed before running that part of the tutorial.

Install missing packages with the following copy and paste block. It is shown as plain code on purpose, not as a runnable notebook cell. Run it manually in the R console only if packages are missing.

``` text
install.packages(c("BiocManager", "ggplot2", "revealjs", "remotes"))

BiocManager::install(c(
  "GenomicRanges",
  "GenomeInfoDb",
  "IRanges",
  "S4Vectors",
  "rtracklayer",
  "BSgenome.Hsapiens.UCSC.hg19",
  "TxDb.Hsapiens.UCSC.hg19.knownGene",
  "Homo.sapiens",
  "AnnotationDbi",
  "GenomicFeatures",
  "Biostrings",
  "Gviz"
))

# Install it from the GenomicsClass GitHub repository used by the course
remotes::install_github("genomicsclass/ERBS")
```

This is setup only. Run it manually only when the package check above shows missing packages.

# ERBS binding peaks

This tutorial uses estrogen receptor binding data from ENCODE style peak files. ERBS data are peak like genomic intervals.

-   a reference genome sequence, `hg19`;
-   a transcript/gene model database, `TxDb.Hsapiens.UCSC.hg19.knownGene`;
-   an organism level ID mapping database, `Homo.sapiens`;
-   interval operations `GRanges`.

# Import ERBS peaks

## Locate and inspect the packaged file

The workflow finds the raw peak like file inside the `ERBS` package. We inspect a few lines before importing.

```{r}
# Find the ERBS data files bundled with the package and store their locations
erbs_files <- dir(system.file("extdata", package = "ERBS"), full.names = TRUE)
erbs_files

# Use the first peak file for the tutorial
f1 <- erbs_files[1]

# Preview the raw interval lines before importing
readLines(f1, n = 4)
```

The file used here is `ENCFF001VEH.narrowPeak`. The preview shows the chromosome the peak is located on, the genomic interval (start, end), and peak metadata (peakID/name, score, strand, signal strength, statistical confidence, adjusted confidence, peak summit offset). 

The next step is to import the raw peak file into `GRanges`

## Import as a computable genomic object

```{r}
library(GenomicRanges) #stores DNA regions
library(GenomeInfoDb) #stores genome information
library(S4Vectors) #stores metadata columns
library(rtracklayer) #imports genomics files
```

These libraries provide genome interval objects, genome labels, metadata columns, and file import.

```{r}
# Import the peak file as genomic ranges
erbs_peaks <- import(f1, format = "bedGraph")
erbs_peaks

# Label these genomic positions as hg19 human genome coordinates
genome(erbs_peaks) <- "hg19"
erbs_peaks
```

Import creates 1,873 ERBS peaks as `GRanges`. Setting the genome to `hg19` keeps later gene, promoter, exon, and sequence operations consistent.

```{r}
# Rename metadata columns for readability
candidate_names <- c(
  "peak_name_raw",
  "score_raw",
  "strand_raw",
  "signalValue",
  "pValue",
  "qValue",
  "peak_offset"
)
names(mcols(erbs_peaks)) <- candidate_names[seq_len(ncol(mcols(erbs_peaks)))]

# Add a simple unique ID for each peak
mcols(erbs_peaks)$peak_id <- paste0("ERBS_peak_", seq_along(erbs_peaks))

head(as.data.frame(mcols(erbs_peaks)))
```

The metadata columns now have readable names such as `signalValue`, `qValue`, and `peak_id`.

## Basic checks

```{r}
# Count peaks
length(erbs_peaks)
```

```{r}
# List chromosomes present in the peak set
seqlevels(erbs_peaks)
```
```{r}
# Confirm the genome build label
genome(erbs_peaks)
```
```{r}
# Summarize peak widths
summary(width(erbs_peaks))
```

```{r}
# Inspect the first few peaks as a table
head(as.data.frame(erbs_peaks))
```

This run has 1,873 peaks. Widths range from 247 bp to 11,894 bp, with a median of 1,310 bp, once the peaks are in `GRanges`

bp (base pairs) = Basic unit of DNA length

## Load hg19 sequence and gene models

```{r}
library(BSgenome.Hsapiens.UCSC.hg19) #Loads the actual human DNA sequence for the hg19 genome
library(TxDb.Hsapiens.UCSC.hg19.knownGene) #loads known gene locations
library(Homo.sapiens) #loads gene name information and ID mappings
library(AnnotationDbi) #annotations
library(GenomicFeatures) #gene features
library(GenomicRanges) #genomi

# Reference genome sequence
Hsapiens

# Transcript database for hg19 known Gene models
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
txdb

# Gene ranges from the transcript database
all_genes <- genes(txdb)

# Keep a stable gene ID column for joins and labels
if (!"gene_id" %in% names(mcols(all_genes))) {
  mcols(all_genes)$gene_id <- names(all_genes)
}

all_genes
```

`BSgenome` supplies the `hg19` sequence and `TxDb` supplies 28,622 gene ranges on the same build

## Genome build consistency

Before overlapping peaks with genes, the chromosome names and genome versions must match. 

```{r}
# Genome label from the reference sequence
head(genome(Hsapiens))

# Genome label on the peak data
unique(genome(erbs_peaks))

# Genome label on the gene models
unique(genome(all_genes))

# Chromosomes shared by peaks and gene models
intersect(seqlevels(erbs_peaks), seqlevels(all_genes))
```

The peaks and gene models share UCSC style chromosome names such as `chr1` and `chr11`. The shared chromosome set confirms these objects can be overlapped safely on `hg19`.

# Annotate the full ERBS peak set

This section uses the whole ERBS dataset. We are not yet zooming to ESRRA. The goal is to create an annotated peak table and whole dataset summaries.

## Prepare annotation ranges

Recall definitions:
- Promoter: a DNA region near the start of a gene where regulatory proteins often bind. Important for turning genes on and controlling gene activity (say 2,000 DNA letters before the gene start and 200 DNA letters after the start).
- Transcription Start Site (TSS): the exact DNA position where the gene reading begins.
- Exons: protein-coding pieces. The parts that actually contribute to making proteins.

```{r}
# Promoter window uses 2 kb upstream and 200 bp downstream
promoter_ranges <- promoters(all_genes, upstream = 2000, downstream = 200)

# One base transcription start site window
tss_ranges <- promoters(all_genes, upstream = 0, downstream = 1)

# Exons grouped by gene, then flattened into one range set
exons_by_gene <- exonsBy(txdb, by = "gene")
all_exons <- unlist(exons_by_gene, use.names = FALSE)

length(all_genes)
length(promoter_ranges)
length(all_exons)
```

The annotation targets are 28,622 genes, 28,622 promoter windows, and 672,344 exon ranges. Promoters are defined here as 2 kb upstream and 200 bp downstream of the TSS

Recall definitions:
- Introns: skipped DNA regions inside genes before protein production
- CDS (coding DNA sequence): the part that determines the protein sequence
- UTR (Untranslated Region): part of the gene transcript not converted into protein. Help regulate stability, translation, gene control (5' UTR - located before CDS, 3' UTR - located after CDS)

```{r}
# Introns are grouped by transcript first then flattened
introns_by_tx <- intronsByTranscript(txdb, use.names = FALSE)
all_introns <- unlist(introns_by_tx, use.names = FALSE)

# Coding sequence ranges are available directly from TxDb
all_cds <- cds(txdb)

# UTR ranges are grouped by transcript first then flattened
five_utrs_by_tx <- fiveUTRsByTranscript(txdb, use.names = FALSE)
three_utrs_by_tx <- threeUTRsByTranscript(txdb, use.names = FALSE)
all_five_utrs <- unlist(five_utrs_by_tx, use.names = FALSE)
all_three_utrs <- unlist(three_utrs_by_tx, use.names = FALSE)

structural_range_counts <- data.frame(
  feature = c("intron", "CDS", "five prime UTR", "three prime UTR"),
  ranges = c(length(all_introns), length(all_cds), length(all_five_utrs), length(all_three_utrs))
)

knitr::kable(structural_range_counts)
```

These extra structural ranges let the same ERBS peaks be annotated beyond promoter, exon, and gene body labels

## Add nearest gene and TSS annotations

```{r}
# Start from the imported peaks
annotated_peaks <- erbs_peaks

length(annotated_peaks)
```

The working annotation object starts with all 1,873 imported ERBS peaks.

Now, we find the closest gene to each ERBS peak and how far away that gene is.

```{r}
# Find the nearest gene range for each peak
nearest_gene_idx <- nearest(annotated_peaks, all_genes, ignore.strand = TRUE)
nearest_gene_id <- rep(NA_character_, length(annotated_peaks))
has_nearest <- !is.na(nearest_gene_idx)
nearest_gene_id[has_nearest] <- all_genes$gene_id[nearest_gene_idx[has_nearest]]

# Measure distance to the nearest gene
nearest_distance <- rep(NA_integer_, length(annotated_peaks))
dtn <- distanceToNearest(annotated_peaks, all_genes, ignore.strand = TRUE)
nearest_distance[queryHits(dtn)] <- mcols(dtn)$distance

# Add nearest gene and distance information to the peaks
mcols(annotated_peaks)$nearest_gene_id <- nearest_gene_id
mcols(annotated_peaks)$distance_to_nearest_gene <- nearest_distance

knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "nearest_gene_id",
  "distance_to_nearest_gene"
)]), 8))
```

Each peak now has a nearest Entrez gene ID and a distance to that gene range. Many peaks overlap a gene body, so their distance is 0 bp. This information helps to guess which genes the binding signals might influence.


Note that a peak can be close to a gene overall while still being far from its important control region. Hence, we also find the nearest TSS to each ERBS peak and how far away it is.

```{r}
# Find the nearest transcription start site for each peak
nearest_tss_idx <- nearest(annotated_peaks, tss_ranges, ignore.strand = TRUE)
nearest_tss_gene_id <- rep(NA_character_, length(annotated_peaks))
has_tss <- !is.na(nearest_tss_idx)
nearest_tss_gene_id[has_tss] <- tss_ranges$gene_id[nearest_tss_idx[has_tss]]

# Measure distance to the nearest transcription start site
tss_distance <- rep(NA_integer_, length(annotated_peaks))
dtt <- distanceToNearest(annotated_peaks, tss_ranges, ignore.strand = TRUE)
tss_distance[queryHits(dtt)] <- mcols(dtt)$distance

mcols(annotated_peaks)$nearest_tss_gene_id <- nearest_tss_gene_id
mcols(annotated_peaks)$distance_to_nearest_tss <- tss_distance

knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "nearest_tss_gene_id",
  "distance_to_nearest_tss"
)]), 8))
```

Conclusion: Distances to TSS are much more variable and informative. Key finding — two genes (7086/TFF1 and 2101/ESRRA) have peaks exactly at the transcription start site (distance = 0). This is the strongest evidence for direct regulation. Other peaks (54, 51341, 101928738) lie in promoter regions (1.6-10 thousand bp), while peaks 139538 and 121233925 are located >140 thousand bp away — likely within introns or enhancers acting at a distance.

The TSS distance gives a promoter oriented view of each peak. It is different from distance to the full gene body because genes can be long and the biologically important control region is usually near the TSS. 

## Add TSS distance bins

Here, we count how many peaks fall into each TSS distance category so as to see whether protein binding events tend to occur near genes.

```{r}
# Define TSS distance ranges and add labels
tss_breaks <- c(-1, 0, 1000, 5000, 10000, 50000, Inf)
tss_labels <- c(
  "0 bp",
  "1 to 1000 bp",
  "1001 to 5000 bp",
  "5001 to 10000 bp",
  "10001 to 50000 bp",
  "over 50000 bp"
)

# Categorize each peak
mcols(annotated_peaks)$tss_distance_bin <- cut(
  mcols(annotated_peaks)$distance_to_nearest_tss,
  breaks = tss_breaks,
  labels = tss_labels
)

# Count peaks in each category
tss_interval_bins <- as.data.frame(table(mcols(annotated_peaks)$tss_distance_bin))
names(tss_interval_bins) <- c("distance_bin", "peaks")

knitr::kable(tss_interval_bins)
```

Conclusion: The majority of ERBS peaks (697 out of 1,873, or 37%) fall exactly at TSS positions (distance = 0 bp), confirming strong enrichment at transcription start sites. 

The bins turn exact TSS distances into a quick promoter proximity summary.

## Add peak summit annotation
Now we calculate the summit. The summit is the strongest point inside a peak when the peak caller reports an offset. An offset is how far the summit is from the start of the peak. 

```{r}
# Read the summit offset
peak_offset <- suppressWarnings(as.integer(mcols(annotated_peaks)$peak_offset))

# Calculate the summit position
summit_position <- start(annotated_peaks) + peak_offset

# Check validity of the summit
valid_summit <- !is.na(summit_position) &
  summit_position >= start(annotated_peaks) &
  summit_position <= end(annotated_peaks)

# Store valid summits
mcols(annotated_peaks)$summit_position <- NA_integer_
mcols(annotated_peaks)$summit_position[valid_summit] <- summit_position[valid_summit]

# Count summits
data.frame(
  peaks = length(annotated_peaks),
  peaks_with_summit = sum(valid_summit)
)
```
The table shows how many peaks have a usable summit position.

Now, we check how far the summit inside each peak is from the nearest TSS.

```{r}
# Create a new genomic object containing only the summit points. Summit is made into a tiny 1bp region
peak_summits <- GRanges(
  seqnames = seqnames(annotated_peaks)[valid_summit],
  ranges = IRanges(start = summit_position[valid_summit], width = 1),
  strand = strand(annotated_peaks)[valid_summit],
  seqinfo = seqinfo(annotated_peaks)
)

# Measure distance from each summit to nearest TSS
summit_tss_distance <- rep(NA_integer_, length(annotated_peaks))
dts <- distanceToNearest(peak_summits, tss_ranges, ignore.strand = TRUE)
summit_tss_distance[which(valid_summit)[queryHits(dts)]] <- mcols(dts)$distance

# Stores summit-to-TSS distance for each peak
mcols(annotated_peaks)$summit_distance_to_nearest_tss <- summit_tss_distance
summary(mcols(annotated_peaks)$summit_distance_to_nearest_tss)
head(as.data.frame(mcols(annotated_peaks)))
```

Summit distance focuses the TSS comparison on one base per peak rather than the full peak interval.

Here, we group the summits by their distance from the TSS and count how many fall into each distance range.

```{r}
mcols(annotated_peaks)$summit_tss_distance_bin <- cut(
  mcols(annotated_peaks)$summit_distance_to_nearest_tss,
  breaks = tss_breaks,
  labels = tss_labels
)

summit_interval_bins <- as.data.frame(table(mcols(annotated_peaks)$summit_tss_distance_bin))
names(summit_interval_bins) <- c("distance_bin", "summits")

knitr::kable(summit_interval_bins)
```

The summit based bins are a stricter view of how close the strongest point of each peak is to a TSS.

## Add gene symbols

Internal gene ID's are now converted into human-readable gene names.

```{r}
# Convert Entrez gene IDs to gene symbols using the Homo.sapiens gene database
nearest_symbol <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = mcols(annotated_peaks)$nearest_gene_id,
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)

# Convert nearest TSS Entrez IDs to gene symbols
nearest_tss_symbol <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = mcols(annotated_peaks)$nearest_tss_gene_id,
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)

# Store results
mcols(annotated_peaks)$nearest_symbol <- unname(nearest_symbol)
mcols(annotated_peaks)$nearest_tss_symbol <- unname(nearest_tss_symbol)

knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "nearest_symbol",
  "nearest_tss_symbol"
)]), 8))
```

When analyzing transcriptional regulation, priority should be given to TSS annotation—particularly when `distance_to_nearest_tss = 0`. This enables the identification of true promoter binding sites, even if another gene is physically located closer.

Gene symbols make the table easier to read, while the Entrez IDs keep the link to the gene model database

## Add feature overlap annotations

We now check if each peak overlaps important parts of the genes.

```{r}
# Mark peaks that touch promoter windows
mcols(annotated_peaks)$overlaps_promoter <- countOverlaps(
  annotated_peaks,
  promoter_ranges,
  ignore.strand = TRUE
) > 0

# Mark peaks that touch exons
mcols(annotated_peaks)$overlaps_exon <- countOverlaps(
  annotated_peaks,
  all_exons,
  ignore.strand = TRUE
) > 0

# Mark peaks that touch gene bodies
mcols(annotated_peaks)$overlaps_gene_body <- countOverlaps(
  annotated_peaks,
  all_genes,
  ignore.strand = TRUE
) > 0

head(as.data.frame(mcols(annotated_peaks)[, c(
  "overlaps_promoter",
  "overlaps_exon",
  "overlaps_gene_body"
)]))
```

These three logical columns mark whether each peak touches promoter, exon, or gene body ranges

We now check if each peak overlaps specific structural parts of the genes.
```{r}
# Mark peaks that touch transcript structural features

# Mark peaks that overlap introns
mcols(annotated_peaks)$overlaps_intron <- countOverlaps(
  annotated_peaks,
  all_introns,
  ignore.strand = TRUE
) > 0

# Mark peaks that overlap CDS
mcols(annotated_peaks)$overlaps_cds <- countOverlaps(
  annotated_peaks,
  all_cds,
  ignore.strand = TRUE
) > 0

# Mark peaks that overlap 5' untranslated region
mcols(annotated_peaks)$overlaps_five_utr <- countOverlaps(
  annotated_peaks,
  all_five_utrs,
  ignore.strand = TRUE
) > 0

# Mark peaks that overlap 3' untranslated region
mcols(annotated_peaks)$overlaps_three_utr <- countOverlaps(
  annotated_peaks,
  all_three_utrs,
  ignore.strand = TRUE
) > 0
```

The structural overlap columns separate coding and untranslated transcript features from broader exon and gene body labels

```{r}
knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "overlaps_intron",
  "overlaps_cds",
  "overlaps_five_utr",
  "overlaps_three_utr"
)]), 8))
```

This preview shows the added structural feature flags for the first few peaks.

Now, we check how many peaks overlap each genomic feature to gain more insights into where binding signals most often occur.

```{r}
overlap_counts <- data.frame(
  feature = c(
    "promoter",
    "exon",
    "intron",
    "gene body",
    "CDS",
    "five prime UTR",
    "three prime UTR"
  ),
  peaks = c(
    sum(mcols(annotated_peaks)$overlaps_promoter, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_exon, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_intron, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_gene_body, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_cds, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_five_utr, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_three_utr, na.rm = TRUE)
  )
)

knitr::kable(overlap_counts)
```

The overlap counts are not mutually exclusive because genomic features can overlap.

## Add promoter linked genes

We now find the exact promoter overlaps and record which peak overlaps which gene's promoter.

```{r}
# Find every case where an ERBS peak touches a promoter region
promoter_hits <- findOverlaps(
  annotated_peaks,
  promoter_ranges,
  ignore.strand = TRUE
)

# Create a table with row number of the peak that overlapped promoters, the peak ID and the gene ID whose promoter is overlapped
promoter_links <- data.frame(
  peak_index = queryHits(promoter_hits),
  peak_id = mcols(annotated_peaks)$peak_id[queryHits(promoter_hits)],
  promoter_gene_id = promoter_ranges$gene_id[subjectHits(promoter_hits)]
)

# Convert numeric gene IDs into readable gene symbols
promoter_symbol_lookup <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = unique(promoter_links$promoter_gene_id),
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)
```

This cell builds one row for each peak to promoter gene link.

```{r}
promoter_links$promoter_symbol <- unname(
  promoter_symbol_lookup[promoter_links$promoter_gene_id]
)

knitr::kable(head(promoter_links, 12))
```

The table shows individual promoter gene links, so repeated peak IDs mean that one peak links to more than one promoter gene.

Now, we take the promoter-overlap results and add them back as new annotation columns to the main peak table. So for each ERBS peak, we record which gene promoter it overlaps.

```{r}
# Create a clean list of unique gene IDs
collapse_unique <- function(x) {
  paste(sort(unique(x[!is.na(x) & x != ""])), collapse = ", ")
}

# Group promoter hits by peak
id_by_peak <- split(promoter_links$promoter_gene_id, promoter_links$peak_index)
symbol_by_peak <- split(promoter_links$promoter_symbol, promoter_links$peak_index)

# Fill in peaks that actaully overlap promoters
promoter_gene_ids <- rep("", length(annotated_peaks))
promoter_symbols <- rep("", length(annotated_peaks))
promoter_gene_ids[as.integer(names(id_by_peak))] <- vapply(id_by_peak, collapse_unique, character(1))
promoter_symbols[as.integer(names(symbol_by_peak))] <- vapply(symbol_by_peak, collapse_unique, character(1))

# Store the results in annotated peaks
mcols(annotated_peaks)$promoter_gene_ids <- promoter_gene_ids
mcols(annotated_peaks)$promoter_symbols <- promoter_symbols
mcols(annotated_peaks)$promoter_gene_count <- tabulate(
  promoter_links$peak_index,
  nbins = length(annotated_peaks)
)
```

The promoter gene lists are now stored on the main annotated peak object

```{r}
promoter_peak_table <- as.data.frame(mcols(annotated_peaks)[
  mcols(annotated_peaks)$promoter_gene_count > 0,
  c("peak_id", "promoter_symbols", "promoter_gene_count")
])

knitr::kable(head(promoter_peak_table, 10))
```

This preview is peak centered, with all promoter linked symbols collapsed into one row per peak

## Add context labels

Initially, every peak is labeled "distal/intergenic". If a peak overlaps a gene, it becomes "gene body". If a peak overlaps an exon, it upgrades to "exon". If it overlaps a promoter, it becomes a "promoter (and an exon)"

```{r}
# Assign one context label per peak
primary_context <- rep("distal/intergenic", length(annotated_peaks))
primary_context[mcols(annotated_peaks)$overlaps_gene_body] <- "gene body"
primary_context[mcols(annotated_peaks)$overlaps_exon] <- "exon"
primary_context[mcols(annotated_peaks)$overlaps_promoter] <- "promoter"
mcols(annotated_peaks)$primary_context <- factor(
  primary_context,
  levels = c("promoter", "exon", "gene body", "distal/intergenic")
)

annotated_peaks

#head(as.data.frame(mcols(annotated_peaks)))
```
```{r}
knitr::kable(
  head(as.data.frame(mcols(annotated_peaks)[, c(
    "peak_id",
    "nearest_symbol",
    "primary_context",
    "distance_to_nearest_tss"
  )]), 10)
)
```

The full `GRanges` now carries nearest gene, nearest TSS, overlap, and primary context metadata for every peak

```{r}
detailed_context <- rep("distal/intergenic", length(annotated_peaks))
detailed_context[mcols(annotated_peaks)$overlaps_gene_body] <- "gene body"
detailed_context[mcols(annotated_peaks)$overlaps_intron] <- "intron"
detailed_context[mcols(annotated_peaks)$overlaps_exon] <- "exon"
detailed_context[mcols(annotated_peaks)$overlaps_cds] <- "CDS"
detailed_context[mcols(annotated_peaks)$overlaps_three_utr] <- "three prime UTR"
detailed_context[mcols(annotated_peaks)$overlaps_five_utr] <- "five prime UTR"
detailed_context[mcols(annotated_peaks)$overlaps_promoter] <- "promoter"

mcols(annotated_peaks)$detailed_context <- factor(
  detailed_context,
  levels = c(
    "promoter",
    "five prime UTR",
    "three prime UTR",
    "CDS",
    "exon",
    "intron",
    "gene body",
    "distal/intergenic"
  )
)

head(as.data.frame(mcols(annotated_peaks)))
```

The detailed context label keeps one readable category per peak while using the new intron, CDS, and UTR overlap fields

## Add GC content from the reference genome

Now, we extract the actual DNA sequence under each peak and measure how much of the sequence is made of G and C bases, storing that as the peak's GC content.

```{r}
# Load DNA sequence tools
library(Biostrings)

# Extract the DNA sequence under each peak
peak_sequences <- getSeq(Hsapiens, annotated_peaks)

# Compute the GC fraction of each peak sequence
mcols(annotated_peaks)$gc_content <- letterFrequency(
  peak_sequences,
  letters = "GC",
  as.prob = TRUE
)[, 1]

summary(mcols(annotated_peaks)$gc_content)
```

GC content is computed from the `hg19` sequence under each peak, median is about 0.55

## Build an annotated peak table

```{r}
# Convert ranges to a regular table for display
peak_table <- as.data.frame(annotated_peaks)

dim(peak_table)

peak_table
```

The table has one row per peak and one column for each coordinate, signal, and annotation field

```{r}
coordinate_columns <- c("seqnames", "start", "end", "width")
signal_columns <- c("signalValue", "qValue", "gc_content")
nearest_columns <- c("nearest_gene_id", "nearest_symbol", "nearest_tss_symbol")
distance_columns <- c("distance_to_nearest_gene", "distance_to_nearest_tss")
summit_columns <- c("summit_position", "summit_distance_to_nearest_tss")
promoter_columns <- c("promoter_symbols", "promoter_gene_count")
context_columns <- c("primary_context", "detailed_context")

columns_to_show <- intersect(
  c(
    coordinate_columns,
    signal_columns,
    nearest_columns,
    distance_columns,
    "tss_distance_bin",
    summit_columns,
    "summit_tss_distance_bin",
    promoter_columns,
    context_columns
  ),
  names(peak_table)
)
```

The display table keeps the most useful teaching columns without printing every overlap flag at once

```{r}
# Show the first annotated peaks
knitr::kable(head(peak_table[, columns_to_show], 12))
```

Each row is one ERBS peak with added biological context, shows the move from raw coordinates to annotated genomic intervals.

# Whole dataset annotation summaries

Before zooming to one locus, we summarize the full ERBS peak set

## Chromosome distribution

```{r}
# Count peaks on each chromosome
chr_counts <- sort(table(as.character(seqnames(annotated_peaks))), decreasing = TRUE)
knitr::kable(head(data.frame(chromosome = names(chr_counts), peaks = as.integer(chr_counts)), 12))

# Plot the distribution across chromosomes
barplot(
  chr_counts,
  las = 2,
  col = "steelblue",
  main = "ERBS peaks by chromosome",
  ylab = "Number of peaks"
)
```

Peaks are genome wide. The largest counts are `chr1` with 200 peaks, `chr19` with 183, and `chr16` with 121

## Peak width distribution

```{r}
# Draw the distribution of peak sizes
hist(
  width(annotated_peaks),
  breaks = 40,
  col = "gray70",
  border = "white",
  main = "Peak width distribution",
  xlab = "Peak width in base pairs"
)

# Print numeric width summaries
summary(width(annotated_peaks))
```

Most peaks are compact, median width is 1,310 bp and the mean is about 1,545 bp

## Signal distribution

```{r}
# Draw the peak signal distribution
hist(
  mcols(annotated_peaks)$signalValue,
  breaks = 40,
  col = "tan",
  border = "white",
  main = "ERBS signal value distribution",
  xlab = "Signal value"
)

# Print numeric signal summaries
summary(mcols(annotated_peaks)$signalValue)
```

Signal is right skewed, most peaks have modest signal, while a small number are much stronger.

## Distance to nearest gene and TSS

```{r}
# Pull nearest gene and nearest TSS distances
gene_distance <- mcols(annotated_peaks)$distance_to_nearest_gene
tss_distance <- mcols(annotated_peaks)$distance_to_nearest_tss

# Compare both distances on a log scale
boxplot(
  list(
    nearest_gene = log10(gene_distance + 1),
    nearest_tss = log10(tss_distance + 1)
  ),
  col = c("darkseagreen3", "lightgoldenrod2"),
  main = "Peak distance to nearest gene and TSS",
  ylab = "log10(distance + 1)"
)

distance_summary <- data.frame(
  measure = c("nearest gene body", "nearest TSS"),
  median_bp = c(median(gene_distance, na.rm = TRUE), median(tss_distance, na.rm = TRUE)),
  mean_bp = c(round(mean(gene_distance, na.rm = TRUE), 1), round(mean(tss_distance, na.rm = TRUE), 1)),
  max_bp = c(max(gene_distance, na.rm = TRUE), max(tss_distance, na.rm = TRUE))
)

knitr::kable(distance_summary)
```

The median nearest gene distance is 0 bp, while the median nearest TSS distance is 4,079 bp. TSS proximity is useful, but it is not proof of direct regulation.

## TSS distance bins

```{r}
tss_bin_counts <- data.frame(
  distance_bin = tss_labels,
  peak_interval = as.integer(table(factor(
    mcols(annotated_peaks)$tss_distance_bin,
    levels = tss_labels
  ))),
  peak_summit = as.integer(table(factor(
    mcols(annotated_peaks)$summit_tss_distance_bin,
    levels = tss_labels
  )))
)

knitr::kable(tss_bin_counts)
```

Interpretation of Results ?

```{r}
tss_bin_matrix <- t(as.matrix(tss_bin_counts[, c("peak_interval", "peak_summit")]))

barplot(
  tss_bin_matrix,
  beside = TRUE,
  names.arg = tss_bin_counts$distance_bin,
  las = 2,
  col = c("steelblue", "darkorange"),
  ylab = "Number of peaks",
  main = "TSS distance bins"
)

legend("topright", legend = rownames(tss_bin_matrix), fill = c("steelblue", "darkorange"))
```

Interpretation ?

## Genomic context categories

```{r}
# Count one context label per peak
primary_context_counts <- as.data.frame(table(mcols(annotated_peaks)$primary_context))
names(primary_context_counts) <- c("primary_context", "peaks")

knitr::kable(primary_context_counts)
```

The strict labels assign each peak to one primary genomic context.

```{r}
detailed_context_counts <- as.data.frame(table(mcols(annotated_peaks)$detailed_context))
names(detailed_context_counts) <- c("detailed_context", "peaks")

knitr::kable(detailed_context_counts)
```

The detailed table separates intron, CDS, and UTR categories where the transcript annotation supports them.

```{r}
# Also show overlapping flags for transparency
overlap_context <- data.frame(
  context = c(
    "promoter overlap",
    "exon overlap",
    "intron overlap",
    "gene body overlap",
    "CDS overlap",
    "five prime UTR overlap",
    "three prime UTR overlap"
  ),
  peaks = c(
    sum(mcols(annotated_peaks)$overlaps_promoter, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_exon, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_intron, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_gene_body, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_cds, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_five_utr, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_three_utr, na.rm = TRUE)
  )
)

knitr::kable(overlap_context)
```

Interpretation ?

```{r}
# Plot the context counts
barplot(
  primary_context_counts$peaks,
  names.arg = primary_context_counts$primary_context,
  las = 2,
  col = c("darkorange", "dodgerblue3", "seagreen4", "gray60"),
  main = "Primary genomic context per peak",
  ylab = "Number of peaks"
)
```

The plot shows 775 promoter peaks, 308 exon peaks, 411 gene body peaks, and 379 distal or intergenic peaks

## Top nearest genes

```{r}
# Count how often each gene symbol is nearest to a peak
nearest_symbol_counts <- sort(table(mcols(annotated_peaks)$nearest_symbol), decreasing = TRUE)
nearest_symbol_counts <- nearest_symbol_counts[names(nearest_symbol_counts) != ""]

# Show the most frequent nearest genes
knitr::kable(head(data.frame(symbol = names(nearest_symbol_counts), peaks = as.integer(nearest_symbol_counts)), 15))
```

No single nearest gene dominates the peak set

# ESRRA visualization locus

The workflow now moves from annotation to genome visualization by focusing on the `ESRRA` neighborhood. We use the same ERBS peak data but zoom into a region small enough to draw as tracks.

## ESRRA identifier

```{r}
library(Homo.sapiens)
library(AnnotationDbi)

# Look up ESRRA in the organism annotation database
eid <- AnnotationDbi::select(
  Homo.sapiens,
  keys = "ESRRA",
  keytype = "SYMBOL",
  columns = "ENTREZID"
)
eid
```

`ESRRA` maps to Entrez ID `2101`, which is the identifier used to retrieve the locus from `TxDb`

## Define the ESRRA gene body and neighborhood

```{r}
esrra_id <- "2101"

# Get the ESRRA gene range
esrra_addr <- genes(txdb, filter = list(gene_id = esrra_id))
esrra_addr

# Expand the view by 500 kb around ESRRA
esrra_neigh <- subsetByOverlaps(all_genes, esrra_addr + 500000)

# Add readable gene symbols to the nearby genes
esrra_neigh$symbol <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = esrra_neigh$gene_id,
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)

knitr::kable(esrra_neigh)
```

In this `hg19` annotation, `ESRRA` is on `chr11` at `64,072,969-64,084,215`. The 500 kb neighborhood contains 27 gene ranges

## Subset annotated ERBS peaks to the ESRRA neighborhood

This view keeps the visualization connected to the same `annotated_peaks` object created earlier in the annotation workflow

```{r}
# Keep annotated peaks that overlap the ESRRA neighborhood
annotated_esrra_window_peaks <- subsetByOverlaps(annotated_peaks, range(esrra_neigh))
annotated_esrra_window_peaks

# Show annotation fields carried into the plot
annotated_esrra_table <- as.data.frame(annotated_esrra_window_peaks)
display_cols <- intersect(
  c(
    "peak_id",
    "seqnames",
    "start",
    "end",
    "width",
    "signalValue",
    "nearest_symbol",
    "nearest_tss_symbol",
    "distance_to_nearest_tss",
    "primary_context",
    "detailed_context"
  ),
  names(annotated_esrra_table)
)
knitr::kable(annotated_esrra_table[, display_cols])
```

The ESRRA neighborhood contains 6 annotated ERBS peaks. Interpretation?

At this point, we have these objects.

-   the annotated ERBS peaks from the tutorial workflow
-   a focused ESRRA genomic neighborhood
-   gene models for the neighborhood
-   peak metadata to plot as track signal

# Gviz visualization of ERBS peaks and ESRRA gene context

Genome visualization asks a different question from whole dataset summaries. Instead of asking "what do all peaks tend to overlap?", it asks, "what evidence do I see at this specific genomic location?"

## Gene view

```{r}
library(Gviz)

# Create a gene model track for nearby genes
gene_track <- GeneRegionTrack(
  esrra_neigh,
  showId = TRUE,
  name = "genes near ESRRA",
  geneSymbol = TRUE
)

plotTracks(gene_track)
```

Interpretation ?

## Plot window

```{r}
library(Gviz)

# Define the plotted genomic window
plot_region <- range(esrra_neigh)
plot_chr <- as.character(seqnames(plot_region))[1]
plot_from <- start(plot_region)
plot_to <- end(plot_region)

data.frame(chromosome = plot_chr, from = plot_from, to = plot_to)
```

The plotting window is the ESRRA neighborhood used earlier, so the visual view stays connected to the same annotation object


Now, we create two helper tracks for the genome visualization. A chromosome overview and a coordinate ruler. 

```{r}
# Add chromosome and coordinate tracks
idx_track <- IdeogramTrack(genome = "hg19", chromosome = plot_chr)
axis_track <- GenomeAxisTrack()

class(idx_track)
class(axis_track)
```
The chromosome overview track displays the whole chromosome and highlights where the ESRRA region sits on it while the genomic coordinate ruler shows DNA positions along the chromosome.

## Gviz with the annotated ERBS peaks

Now, we create a visualization track for ERBS peaks near ESRRA, using signal strengths when available so the plot can show how strong the binding signals are across the genomic region.

```{r}
# Use signal values from the annotated peaks when they are available
annotated_signal_track <- annotated_esrra_window_peaks
if ("signalValue" %in% names(mcols(annotated_signal_track))) {
  mcols(annotated_signal_track) <- S4Vectors::DataFrame(signalValue = mcols(annotated_signal_track)$signalValue)
  annotated_peak_track <- DataTrack(
    annotated_signal_track,
    data = "signalValue",
    type = c("h", "p"),
    name = "annotated ERBS signal",
    genome = "hg19",
    chromosome = plot_chr
  )
} else {
  annotated_peak_track <- AnnotationTrack(
    annotated_esrra_window_peaks,
    name = "annotated ERBS peaks",
    genome = "hg19",
    chromosome = plot_chr
  )
}

class(annotated_peak_track)
```

Now, we create the gene layer for the final genome visualization. This tells Gviz to draw the genes located near ESSRA.

```{r}
# Add nearby gene models
gene_track <- GeneRegionTrack(
  esrra_neigh,
  showId = TRUE,
  name = "genes near ESRRA",
  geneSymbol = TRUE,
  genome = "hg19",
  chromosome = plot_chr
)

class(gene_track)
```

The gene track draws nearby transcript models, giving the ERBS signal a genomic annotation context.

```{r}
# Draw the annotated peaks with gene context
plotTracks(
  list(idx_track, axis_track, annotated_peak_track, gene_track),
  from = plot_from,
  to = plot_to,
  chromosome = plot_chr,
  background.title = "gray90",
  col.title = "black",
  cex.title = 0.8
)
```

This combined plot places the 6 annotated ERBS peaks above the ESRRA neighborhood gene models in a genome track view.

Interpretation ?

Questions

-   Which peaks fall near ESRRA?

-   Are the strongest peaks inside a gene body, near a promoter, or distal?

-   How close are the peaks to neighboring genes?

-   What evidence would be needed to claim direct regulation?

-   Would this view change if the coordinates were interpreted as hg38 instead of hg19?
