---
title: "Mapping Sequencing Data to Genome Annotation and Visualization"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 5,
  dpi = 120
)

pkg_ok <- function(pkg) {
  requireNamespace(pkg, quietly = TRUE)
}

need <- function(pkgs) {
  all(vapply(pkgs, pkg_ok, logical(1)))
}

missing_pkgs <- function(pkgs) {
  pkgs[!vapply(pkgs, pkg_ok, logical(1))]
}

erbs_peaks <- NULL
annotated_peaks <- NULL
gm12878_peaks <- NULL
txdb <- NULL
all_genes <- NULL
all_exons <- NULL
promoter_ranges <- NULL
tss_ranges <- NULL
esrra_neigh <- NULL
annotated_esrra_window_peaks <- NULL
esrra_window_peaks <- NULL
```

# Tutorial

This is a practical ERBS tutorial. It starts with the ERBS peak file, turns it into a GRanges object, annotates every peak against the human `hg19` genome, and then visualizes the ESRRA locus with Gviz.

This notebook is written to be runnable when the relevant Bioconductor packages are installed.

```{r}
core_packages <- c(
  "knitr",
  "BiocManager",
  "GenomicRanges",
  "GenomeInfoDb",
  "IRanges",
  "S4Vectors",
  "rtracklayer",
  "ERBS",
  "BSgenome",
  "BSgenome.Hsapiens.UCSC.hg19",
  "TxDb.Hsapiens.UCSC.hg19.knownGene",
  "Homo.sapiens",
  "AnnotationDbi",
  "GenomicFeatures",
  "Biostrings",
  "Gviz",
  "ggplot2"
)
```

```{r}
package_status <- data.frame(
  package = core_packages,
  installed = vapply(core_packages, pkg_ok, logical(1)),
  row.names = NULL
)

knitr::kable(package_status)
```

TRUE means the package is available. Any FALSE row marks a package that should be installed before running that part of the tutorial.

Install missing packages with the following copy and paste block. It is shown as plain code on purpose, not as a runnable notebook cell. Run it manually in the R console only if packages are missing.

``` text
install.packages(c("BiocManager", "ggplot2", "revealjs", "remotes"))

BiocManager::install(c(
  "GenomicRanges",
  "GenomeInfoDb",
  "IRanges",
  "S4Vectors",
  "rtracklayer",
  "BSgenome.Hsapiens.UCSC.hg19",
  "TxDb.Hsapiens.UCSC.hg19.knownGene",
  "Homo.sapiens",
  "AnnotationDbi",
  "GenomicFeatures",
  "Biostrings",
  "Gviz"
))

# Install it from the GenomicsClass GitHub repository used by the course
remotes::install_github("genomicsclass/ERBS")
```

This is setup only. Run it manually only when the package check above shows missing packages.

# ERBS binding peaks

This tutorial uses estrogen receptor binding data from ENCODE style peak files. ERBS data are peak like genomic intervals.

-   a reference genome sequence, `hg19`;
-   a transcript/gene model database, `TxDb.Hsapiens.UCSC.hg19.knownGene`;
-   an organism level ID mapping database, `Homo.sapiens`;
-   interval operations `GRanges`.

# Import ERBS peaks

## Locate and inspect the packaged file

The workflow finds the raw peak like file inside the `ERBS` package. We inspect a few lines before importing.

```{r}
# Find the ERBS data files bundled with the package
erbs_files <- dir(system.file("extdata", package = "ERBS"), full.names = TRUE)
erbs_files

# Use the first peak file for the tutorial
f1 <- erbs_files[1]

# Preview the raw interval lines before importing
readLines(f1, n = 4)
```

The file used here is `ENCFF001VEH.narrowPeak`. The preview shows chromosome, start, end, and peak metadata, so the next step is import into `GRanges`

## Import as a computable genomic object

```{r}
library(GenomicRanges)
library(GenomeInfoDb)
library(S4Vectors)
library(rtracklayer)
```

These libraries provide genome interval objects, genome labels, metadata columns, and file import.

```{r}
# Import the peak file as genomic ranges
erbs_peaks <- import(f1, format = "bedGraph")
erbs_peaks

# Mark the coordinate system used by these peaks
genome(erbs_peaks) <- "hg19"
erbs_peaks
```

Import creates 1,873 ERBS peaks as `GRanges`. Setting the genome to `hg19` keeps later gene, promoter, exon, and sequence operations consistent.

```{r}
# Rename metadata columns for readability
candidate_names <- c(
  "peak_name_raw",
  "score_raw",
  "strand_raw",
  "signalValue",
  "pValue",
  "qValue",
  "peak_offset"
)
names(mcols(erbs_peaks)) <- candidate_names[seq_len(ncol(mcols(erbs_peaks)))]

# Add a simple ID for each peak
mcols(erbs_peaks)$peak_id <- paste0("ERBS_peak_", seq_along(erbs_peaks))

head(as.data.frame(mcols(erbs_peaks)))
```

The metadata columns now have readable names such as `signalValue`, `qValue`, and `peak_id`.

## Basic checks

```{r}
# Count peaks
length(erbs_peaks)

# List chromosomes present in the peak set
seqlevels(erbs_peaks)

# Confirm the genome build label
genome(erbs_peaks)

# Summarize peak widths
summary(width(erbs_peaks))

# Inspect the first few peaks as a table
head(as.data.frame(erbs_peaks))
```

This run has 1,873 peaks. Widths range from 247 bp to 11,894 bp, with a median of 1,310 bp. Once the peaks are in `GRanges`

## Load hg19 sequence and gene models

```{r}
library(BSgenome.Hsapiens.UCSC.hg19)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(Homo.sapiens)
library(AnnotationDbi)
library(GenomicFeatures)
library(GenomicRanges)

# Reference genome sequence
Hsapiens

# Transcript database for hg19 knownGene models
txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene
txdb

# Gene ranges from the transcript database
all_genes <- genes(txdb)

# Keep a stable gene ID column for joins and labels
if (!"gene_id" %in% names(mcols(all_genes))) {
  mcols(all_genes)$gene_id <- names(all_genes)
}

all_genes
```

`BSgenome` supplies the `hg19` sequence and `TxDb` supplies 28,622 gene ranges on the same build

## Genome build consistency

```{r}
# Genome label from the reference sequence
head(genome(Hsapiens))

# Genome label on the peak data
unique(genome(erbs_peaks))

# Genome label on the gene models
unique(genome(all_genes))

# Chromosomes shared by peaks and gene models
intersect(seqlevels(erbs_peaks), seqlevels(all_genes))
```

The peaks and gene models share UCSC style chromosome names such as `chr1` and `chr11`. The shared chromosome set confirms these objects can be overlapped safely on `hg19`.

# Annotate the full ERBS peak set

This section uses the whole ERBS dataset. We are not yet zooming to ESRRA. The goal is to create an annotated peak table and whole dataset summaries.

## Prepare annotation ranges

```{r}
# Promoter window uses 2 kb upstream and 200 bp downstream
promoter_ranges <- promoters(all_genes, upstream = 2000, downstream = 200)

# One base transcription start site window
tss_ranges <- promoters(all_genes, upstream = 0, downstream = 1)

# Exons grouped by gene, then flattened into one range set
exons_by_gene <- exonsBy(txdb, by = "gene")
all_exons <- unlist(exons_by_gene, use.names = FALSE)

length(all_genes)
length(promoter_ranges)
length(all_exons)
```

The annotation targets are 28,622 genes, 28,622 promoter windows, and 672,344 exon ranges, Promoters are defined here as 2 kb upstream and 200 bp downstream of the TSS

```{r}
# Introns are grouped by transcript before flattening
introns_by_tx <- intronsByTranscript(txdb, use.names = FALSE)
all_introns <- unlist(introns_by_tx, use.names = FALSE)

# Coding sequence ranges are available directly from TxDb
all_cds <- cds(txdb)

# UTR ranges are grouped by transcript before flattening
five_utrs_by_tx <- fiveUTRsByTranscript(txdb, use.names = FALSE)
three_utrs_by_tx <- threeUTRsByTranscript(txdb, use.names = FALSE)
all_five_utrs <- unlist(five_utrs_by_tx, use.names = FALSE)
all_three_utrs <- unlist(three_utrs_by_tx, use.names = FALSE)

structural_range_counts <- data.frame(
  feature = c("intron", "CDS", "five prime UTR", "three prime UTR"),
  ranges = c(length(all_introns), length(all_cds), length(all_five_utrs), length(all_three_utrs))
)

knitr::kable(structural_range_counts)
```

These extra structural ranges let the same ERBS peaks be annotated beyond promoter, exon, and gene body labels

## Add nearest gene and TSS annotations

```{r}
# Start from the imported peaks
annotated_peaks <- erbs_peaks

length(annotated_peaks)
```

The working annotation object starts with all 1,873 imported ERBS peaks

```{r}
# Find the nearest gene range for each peak
nearest_gene_idx <- nearest(annotated_peaks, all_genes, ignore.strand = TRUE)
nearest_gene_id <- rep(NA_character_, length(annotated_peaks))
has_nearest <- !is.na(nearest_gene_idx)
nearest_gene_id[has_nearest] <- all_genes$gene_id[nearest_gene_idx[has_nearest]]

# Measure distance to the nearest gene
nearest_distance <- rep(NA_integer_, length(annotated_peaks))
dtn <- distanceToNearest(annotated_peaks, all_genes, ignore.strand = TRUE)
nearest_distance[queryHits(dtn)] <- mcols(dtn)$distance

mcols(annotated_peaks)$nearest_gene_id <- nearest_gene_id
mcols(annotated_peaks)$distance_to_nearest_gene <- nearest_distance

knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "nearest_gene_id",
  "distance_to_nearest_gene"
)]), 8))
```

Each peak now has a nearest Entrez gene ID and a distance to that gene range. Many peaks overlap a gene body, so their distance is 0 bp

```{r}
# Find the nearest transcription start site for each peak
nearest_tss_idx <- nearest(annotated_peaks, tss_ranges, ignore.strand = TRUE)
nearest_tss_gene_id <- rep(NA_character_, length(annotated_peaks))
has_tss <- !is.na(nearest_tss_idx)
nearest_tss_gene_id[has_tss] <- tss_ranges$gene_id[nearest_tss_idx[has_tss]]

# Measure distance to the nearest transcription start site
tss_distance <- rep(NA_integer_, length(annotated_peaks))
dtt <- distanceToNearest(annotated_peaks, tss_ranges, ignore.strand = TRUE)
tss_distance[queryHits(dtt)] <- mcols(dtt)$distance

mcols(annotated_peaks)$nearest_tss_gene_id <- nearest_tss_gene_id
mcols(annotated_peaks)$distance_to_nearest_tss <- tss_distance

knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "nearest_tss_gene_id",
  "distance_to_nearest_tss"
)]), 8))
```

The TSS distance gives a promoter oriented view of each peak. It is different from distance to the full gene body because genes can be long

## Add TSS distance bins

```{r}
tss_breaks <- c(-1, 0, 1000, 5000, 10000, 50000, Inf)
tss_labels <- c(
  "0 bp",
  "1 to 1000 bp",
  "1001 to 5000 bp",
  "5001 to 10000 bp",
  "10001 to 50000 bp",
  "over 50000 bp"
)

mcols(annotated_peaks)$tss_distance_bin <- cut(
  mcols(annotated_peaks)$distance_to_nearest_tss,
  breaks = tss_breaks,
  labels = tss_labels
)

tss_interval_bins <- as.data.frame(table(mcols(annotated_peaks)$tss_distance_bin))
names(tss_interval_bins) <- c("distance_bin", "peaks")

knitr::kable(tss_interval_bins)
```

The bins turn exact TSS distances into a quick promoter proximity summary

## Add peak summit annotation

```{r}
peak_offset <- suppressWarnings(as.integer(mcols(annotated_peaks)$peak_offset))
summit_position <- start(annotated_peaks) + peak_offset
valid_summit <- !is.na(summit_position) &
  summit_position >= start(annotated_peaks) &
  summit_position <= end(annotated_peaks)

mcols(annotated_peaks)$summit_position <- NA_integer_
mcols(annotated_peaks)$summit_position[valid_summit] <- summit_position[valid_summit]

data.frame(
  peaks = length(annotated_peaks),
  peaks_with_summit = sum(valid_summit)
)
```

The summit is the strongest point inside a peak when the peak caller reports an offset. The table shows how many peaks have a usable summit position

```{r}
peak_summits <- GRanges(
  seqnames = seqnames(annotated_peaks)[valid_summit],
  ranges = IRanges(start = summit_position[valid_summit], width = 1),
  strand = strand(annotated_peaks)[valid_summit],
  seqinfo = seqinfo(annotated_peaks)
)

summit_tss_distance <- rep(NA_integer_, length(annotated_peaks))
dts <- distanceToNearest(peak_summits, tss_ranges, ignore.strand = TRUE)
summit_tss_distance[which(valid_summit)[queryHits(dts)]] <- mcols(dts)$distance

mcols(annotated_peaks)$summit_distance_to_nearest_tss <- summit_tss_distance
summary(mcols(annotated_peaks)$summit_distance_to_nearest_tss)
```

Summit distance focuses the TSS comparison on one base per peak rather than the full peak interval

```{r}
mcols(annotated_peaks)$summit_tss_distance_bin <- cut(
  mcols(annotated_peaks)$summit_distance_to_nearest_tss,
  breaks = tss_breaks,
  labels = tss_labels
)

summit_interval_bins <- as.data.frame(table(mcols(annotated_peaks)$summit_tss_distance_bin))
names(summit_interval_bins) <- c("distance_bin", "peaks")

knitr::kable(summit_interval_bins)
```

The summit based bins are a stricter view of how close the strongest point of each peak is to a TSS

## Add gene symbols

```{r}
# Convert Entrez gene IDs to gene symbols
nearest_symbol <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = mcols(annotated_peaks)$nearest_gene_id,
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)

# Convert nearest TSS Entrez IDs to gene symbols
nearest_tss_symbol <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = mcols(annotated_peaks)$nearest_tss_gene_id,
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)

mcols(annotated_peaks)$nearest_symbol <- unname(nearest_symbol)
mcols(annotated_peaks)$nearest_tss_symbol <- unname(nearest_tss_symbol)

knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "nearest_symbol",
  "nearest_tss_symbol"
)]), 8))
```

Gene symbols make the table easier to read, while the Entrez IDs keep the link to the gene model database

## Add feature overlap annotations

```{r}
# Mark peaks that touch promoter windows
mcols(annotated_peaks)$overlaps_promoter <- countOverlaps(
  annotated_peaks,
  promoter_ranges,
  ignore.strand = TRUE
) > 0
# Mark peaks that touch exons
mcols(annotated_peaks)$overlaps_exon <- countOverlaps(
  annotated_peaks,
  all_exons,
  ignore.strand = TRUE
) > 0
# Mark peaks that touch gene bodies
mcols(annotated_peaks)$overlaps_gene_body <- countOverlaps(
  annotated_peaks,
  all_genes,
  ignore.strand = TRUE
) > 0
head(as.data.frame(mcols(annotated_peaks)[, c(
  "overlaps_promoter",
  "overlaps_exon",
  "overlaps_gene_body"
)]))
```

These three logical columns mark whether each peak touches promoter, exon, or gene body ranges

```{r}
# Mark peaks that touch transcript structural features
mcols(annotated_peaks)$overlaps_intron <- countOverlaps(
  annotated_peaks,
  all_introns,
  ignore.strand = TRUE
) > 0
mcols(annotated_peaks)$overlaps_cds <- countOverlaps(
  annotated_peaks,
  all_cds,
  ignore.strand = TRUE
) > 0
mcols(annotated_peaks)$overlaps_five_utr <- countOverlaps(
  annotated_peaks,
  all_five_utrs,
  ignore.strand = TRUE
) > 0
mcols(annotated_peaks)$overlaps_three_utr <- countOverlaps(
  annotated_peaks,
  all_three_utrs,
  ignore.strand = TRUE
) > 0
```

The structural overlap columns separate coding and untranslated transcript features from broader exon and gene body labels

```{r}
knitr::kable(head(as.data.frame(mcols(annotated_peaks)[, c(
  "overlaps_intron",
  "overlaps_cds",
  "overlaps_five_utr",
  "overlaps_three_utr"
)]), 8))
```

This preview shows the added structural feature flags for the first few peaks

```{r}
overlap_counts <- data.frame(
  feature = c(
    "promoter",
    "exon",
    "intron",
    "gene body",
    "CDS",
    "five prime UTR",
    "three prime UTR"
  ),
  peaks = c(
    sum(mcols(annotated_peaks)$overlaps_promoter, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_exon, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_intron, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_gene_body, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_cds, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_five_utr, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_three_utr, na.rm = TRUE)
  )
)

knitr::kable(overlap_counts)
```

The overlap counts are not mutually exclusive because genomic features can overlap

## Add promoter linked genes

```{r}
promoter_hits <- findOverlaps(
  annotated_peaks,
  promoter_ranges,
  ignore.strand = TRUE
)

promoter_links <- data.frame(
  peak_index = queryHits(promoter_hits),
  peak_id = mcols(annotated_peaks)$peak_id[queryHits(promoter_hits)],
  promoter_gene_id = promoter_ranges$gene_id[subjectHits(promoter_hits)]
)

promoter_symbol_lookup <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = unique(promoter_links$promoter_gene_id),
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)
```

This cell builds one row for each peak to promoter gene link

```{r}
promoter_links$promoter_symbol <- unname(
  promoter_symbol_lookup[promoter_links$promoter_gene_id]
)

knitr::kable(head(promoter_links, 12))
```

The table shows individual promoter gene links, so repeated peak IDs mean that one peak links to more than one promoter gene

```{r}
collapse_unique <- function(x) {
  paste(sort(unique(x[!is.na(x) & x != ""])), collapse = ", ")
}

id_by_peak <- split(promoter_links$promoter_gene_id, promoter_links$peak_index)
symbol_by_peak <- split(promoter_links$promoter_symbol, promoter_links$peak_index)

promoter_gene_ids <- rep("", length(annotated_peaks))
promoter_symbols <- rep("", length(annotated_peaks))
promoter_gene_ids[as.integer(names(id_by_peak))] <- vapply(id_by_peak, collapse_unique, character(1))
promoter_symbols[as.integer(names(symbol_by_peak))] <- vapply(symbol_by_peak, collapse_unique, character(1))

mcols(annotated_peaks)$promoter_gene_ids <- promoter_gene_ids
mcols(annotated_peaks)$promoter_symbols <- promoter_symbols
mcols(annotated_peaks)$promoter_gene_count <- tabulate(
  promoter_links$peak_index,
  nbins = length(annotated_peaks)
)
```

The promoter gene lists are now stored on the main annotated peak object

```{r}
promoter_peak_table <- as.data.frame(mcols(annotated_peaks)[
  mcols(annotated_peaks)$promoter_gene_count > 0,
  c("peak_id", "promoter_symbols", "promoter_gene_count")
])

knitr::kable(head(promoter_peak_table, 10))
```

This preview is peak centered, with all promoter linked symbols collapsed into one row per peak

## Add context labels

```{r}
# Assign one context label per peak
primary_context <- rep("distal/intergenic", length(annotated_peaks))
primary_context[mcols(annotated_peaks)$overlaps_gene_body] <- "gene body"
primary_context[mcols(annotated_peaks)$overlaps_exon] <- "exon"
primary_context[mcols(annotated_peaks)$overlaps_promoter] <- "promoter"
mcols(annotated_peaks)$primary_context <- factor(
  primary_context,
  levels = c("promoter", "exon", "gene body", "distal/intergenic")
)

annotated_peaks
```

The full `GRanges` now carries nearest gene, nearest TSS, overlap, and primary context metadata for every peak

```{r}
detailed_context <- rep("distal/intergenic", length(annotated_peaks))
detailed_context[mcols(annotated_peaks)$overlaps_gene_body] <- "gene body"
detailed_context[mcols(annotated_peaks)$overlaps_intron] <- "intron"
detailed_context[mcols(annotated_peaks)$overlaps_exon] <- "exon"
detailed_context[mcols(annotated_peaks)$overlaps_cds] <- "CDS"
detailed_context[mcols(annotated_peaks)$overlaps_three_utr] <- "three prime UTR"
detailed_context[mcols(annotated_peaks)$overlaps_five_utr] <- "five prime UTR"
detailed_context[mcols(annotated_peaks)$overlaps_promoter] <- "promoter"

mcols(annotated_peaks)$detailed_context <- factor(
  detailed_context,
  levels = c(
    "promoter",
    "five prime UTR",
    "three prime UTR",
    "CDS",
    "exon",
    "intron",
    "gene body",
    "distal/intergenic"
  )
)
```

The detailed context label keeps one readable category per peak while using the new intron, CDS, and UTR overlap fields

## Add GC content from the reference genome

```{r}
library(Biostrings)

# Extract the DNA sequence under each peak
peak_sequences <- getSeq(Hsapiens, annotated_peaks)

# Compute the GC fraction of each peak sequence
mcols(annotated_peaks)$gc_content <- letterFrequency(
  peak_sequences,
  letters = "GC",
  as.prob = TRUE
)[, 1]

summary(mcols(annotated_peaks)$gc_content)
```

GC content is computed from the `hg19` sequence under each peak, median is about 0.55

## Build an annotated peak table

```{r}
# Convert ranges to a regular table for display
peak_table <- as.data.frame(annotated_peaks)
dim(peak_table)
```

The table has one row per peak and one column for each coordinate, signal, and annotation field

```{r}
coordinate_columns <- c("seqnames", "start", "end", "width")
signal_columns <- c("signalValue", "qValue", "gc_content")
nearest_columns <- c("nearest_gene_id", "nearest_symbol", "nearest_tss_symbol")
distance_columns <- c("distance_to_nearest_gene", "distance_to_nearest_tss")
summit_columns <- c("summit_position", "summit_distance_to_nearest_tss")
promoter_columns <- c("promoter_symbols", "promoter_gene_count")
context_columns <- c("primary_context", "detailed_context")

columns_to_show <- intersect(
  c(
    coordinate_columns,
    signal_columns,
    nearest_columns,
    distance_columns,
    "tss_distance_bin",
    summit_columns,
    "summit_tss_distance_bin",
    promoter_columns,
    context_columns
  ),
  names(peak_table)
)
```

The display table keeps the most useful teaching columns without printing every overlap flag at once

```{r}
# Show the first annotated peaks
knitr::kable(head(peak_table[, columns_to_show], 12))
```

Each row is one ERBS peak with added biological context, shows the move from raw coordinates to annotated genomic intervals.

# Whole dataset annotation summaries

Before zooming to one locus, we summarize the full ERBS peak set

## Chromosome distribution

```{r}
# Count peaks on each chromosome
chr_counts <- sort(table(as.character(seqnames(annotated_peaks))), decreasing = TRUE)
knitr::kable(head(data.frame(chromosome = names(chr_counts), peaks = as.integer(chr_counts)), 12))

# Plot the distribution across chromosomes
barplot(
  chr_counts,
  las = 2,
  col = "steelblue",
  main = "ERBS peaks by chromosome",
  ylab = "Number of peaks"
)
```

Peaks are genome wide. The largest counts are `chr1` with 200 peaks, `chr19` with 183, and `chr16` with 121

## Peak width distribution

```{r}
# Draw the distribution of peak sizes
hist(
  width(annotated_peaks),
  breaks = 40,
  col = "gray70",
  border = "white",
  main = "Peak width distribution",
  xlab = "Peak width in base pairs"
)

# Print numeric width summaries
summary(width(annotated_peaks))
```

Most peaks are compact, median width is 1,310 bp and the mean is about 1,545 bp

## Signal distribution

```{r}
# Draw the peak signal distribution
hist(
  mcols(annotated_peaks)$signalValue,
  breaks = 40,
  col = "tan",
  border = "white",
  main = "ERBS signal value distribution",
  xlab = "Signal value"
)

# Print numeric signal summaries
summary(mcols(annotated_peaks)$signalValue)
```

Signal is right skewed, most peaks have modest signal, while a small number are much stronger.

## Distance to nearest gene and TSS

```{r}
# Pull nearest gene and nearest TSS distances
gene_distance <- mcols(annotated_peaks)$distance_to_nearest_gene
tss_distance <- mcols(annotated_peaks)$distance_to_nearest_tss

# Compare both distances on a log scale
boxplot(
  list(
    nearest_gene = log10(gene_distance + 1),
    nearest_tss = log10(tss_distance + 1)
  ),
  col = c("darkseagreen3", "lightgoldenrod2"),
  main = "Peak distance to nearest gene and TSS",
  ylab = "log10(distance + 1)"
)

distance_summary <- data.frame(
  measure = c("nearest gene body", "nearest TSS"),
  median_bp = c(median(gene_distance, na.rm = TRUE), median(tss_distance, na.rm = TRUE)),
  mean_bp = c(round(mean(gene_distance, na.rm = TRUE), 1), round(mean(tss_distance, na.rm = TRUE), 1)),
  max_bp = c(max(gene_distance, na.rm = TRUE), max(tss_distance, na.rm = TRUE))
)

knitr::kable(distance_summary)
```

The median nearest gene distance is 0 bp, while the median nearest TSS distance is 4,079 bp. TSS proximity is useful, but it is not proof of direct regulation.

## TSS distance bins

```{r}
tss_bin_counts <- data.frame(
  distance_bin = tss_labels,
  peak_interval = as.integer(table(factor(
    mcols(annotated_peaks)$tss_distance_bin,
    levels = tss_labels
  ))),
  peak_summit = as.integer(table(factor(
    mcols(annotated_peaks)$summit_tss_distance_bin,
    levels = tss_labels
  )))
)

knitr::kable(tss_bin_counts)
```

Interpretation of Results ?

```{r}
tss_bin_matrix <- t(as.matrix(tss_bin_counts[, c("peak_interval", "peak_summit")]))

barplot(
  tss_bin_matrix,
  beside = TRUE,
  names.arg = tss_bin_counts$distance_bin,
  las = 2,
  col = c("steelblue", "darkorange"),
  ylab = "Number of peaks",
  main = "TSS distance bins"
)

legend("topright", legend = rownames(tss_bin_matrix), fill = c("steelblue", "darkorange"))
```

Interpretation ?

## Genomic context categories

```{r}
# Count one context label per peak
primary_context_counts <- as.data.frame(table(mcols(annotated_peaks)$primary_context))
names(primary_context_counts) <- c("primary_context", "peaks")

knitr::kable(primary_context_counts)
```

The strict labels assign each peak to one primary genomic context.

```{r}
detailed_context_counts <- as.data.frame(table(mcols(annotated_peaks)$detailed_context))
names(detailed_context_counts) <- c("detailed_context", "peaks")

knitr::kable(detailed_context_counts)
```

The detailed table separates intron, CDS, and UTR categories where the transcript annotation supports them.

```{r}
# Also show overlapping flags for transparency
overlap_context <- data.frame(
  context = c(
    "promoter overlap",
    "exon overlap",
    "intron overlap",
    "gene body overlap",
    "CDS overlap",
    "five prime UTR overlap",
    "three prime UTR overlap"
  ),
  peaks = c(
    sum(mcols(annotated_peaks)$overlaps_promoter, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_exon, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_intron, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_gene_body, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_cds, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_five_utr, na.rm = TRUE),
    sum(mcols(annotated_peaks)$overlaps_three_utr, na.rm = TRUE)
  )
)

knitr::kable(overlap_context)
```

Interpretation ?

```{r}
# Plot the context counts
barplot(
  primary_context_counts$peaks,
  names.arg = primary_context_counts$primary_context,
  las = 2,
  col = c("darkorange", "dodgerblue3", "seagreen4", "gray60"),
  main = "Primary genomic context per peak",
  ylab = "Number of peaks"
)
```

The plot shows 775 promoter peaks, 308 exon peaks, 411 gene body peaks, and 379 distal or intergenic peaks

## Top nearest genes

```{r}
# Count how often each gene symbol is nearest to a peak
nearest_symbol_counts <- sort(table(mcols(annotated_peaks)$nearest_symbol), decreasing = TRUE)
nearest_symbol_counts <- nearest_symbol_counts[names(nearest_symbol_counts) != ""]

# Show the most frequent nearest genes
knitr::kable(head(data.frame(symbol = names(nearest_symbol_counts), peaks = as.integer(nearest_symbol_counts)), 15))
```

No single nearest gene dominates the peak set

# ESRRA visualization locus

The workflow now moves from annotation to genome visualization by focusing on the `ESRRA` neighborhood. We use the same ERBS peak data but zoom into a region small enough to draw as tracks.

## ESRRA identifier

```{r}
library(Homo.sapiens)
library(AnnotationDbi)

# Look up ESRRA in the organism annotation database
eid <- AnnotationDbi::select(
  Homo.sapiens,
  keys = "ESRRA",
  keytype = "SYMBOL",
  columns = "ENTREZID"
)
eid
```

`ESRRA` maps to Entrez ID `2101`, which is the identifier used to retrieve the locus from `TxDb`

## Define the ESRRA gene body and neighborhood

```{r}
esrra_id <- "2101"

# Get the ESRRA gene range
esrra_addr <- genes(txdb, filter = list(gene_id = esrra_id))
esrra_addr

# Expand the view by 500 kb around ESRRA
esrra_neigh <- subsetByOverlaps(all_genes, esrra_addr + 500000)

# Add readable gene symbols to the nearby genes
esrra_neigh$symbol <- AnnotationDbi::mapIds(
  Homo.sapiens,
  keys = esrra_neigh$gene_id,
  keytype = "ENTREZID",
  column = "SYMBOL",
  multiVals = "first"
)

esrra_neigh
```

In this `hg19` annotation, `ESRRA` is on `chr11` at `64,072,969-64,084,215`. The 500 kb neighborhood contains 27 gene ranges

## Subset annotated ERBS peaks to the ESRRA neighborhood

This view keeps the visualization connected to the same `annotated_peaks` object created earlier in the annotation workflow

```{r}
# Keep annotated peaks that overlap the ESRRA neighborhood
annotated_esrra_window_peaks <- subsetByOverlaps(annotated_peaks, range(esrra_neigh))
annotated_esrra_window_peaks

# Show annotation fields carried into the plot
annotated_esrra_table <- as.data.frame(annotated_esrra_window_peaks)
display_cols <- intersect(
  c(
    "peak_id",
    "seqnames",
    "start",
    "end",
    "width",
    "signalValue",
    "nearest_symbol",
    "nearest_tss_symbol",
    "distance_to_nearest_tss",
    "primary_context",
    "detailed_context"
  ),
  names(annotated_esrra_table)
)
knitr::kable(annotated_esrra_table[, display_cols])
```

The ESRRA neighborhood contains 6 annotated ERBS peaks.

## Load GM12878 data for Genome Visualization

```{r}
library(ERBS)

# Load the prepared peak object
data(GM12878)
gm12878_peaks <- GM12878
gm12878_peaks
```

`GM12878` contains 1,873 ERBS peak intervals

## Subset the same peak data to the ESRRA neighborhood

```{r}
# Keep only peaks that overlap the ESRRA neighborhood
esrra_window_peaks <- subsetByOverlaps(gm12878_peaks, range(esrra_neigh))
esrra_window_peaks

# Display the local peaks as a table
knitr::kable(as.data.frame(esrra_window_peaks))
```

The ESRRA neighborhood contains 6 GM12878 peaks.

At this point, we have these objects.

-   the annotated ERBS peaks from the tutorial workflow
-   the official `GM12878` object used by Chapter 14
-   a focused ESRRA genomic neighborhood
-   gene models for the neighborhood
-   peak metadata to plot as track signal

# Gviz visualization of ERBS peaks and ESRRA gene context

Genome visualization asks a different question from whole dataset summaries. Instead of asking "what do all peaks tend to overlap?", we would be looking towards "what evidence do I see at this genomic location?"

## Gene view

```{r}
library(Gviz)

# Create a gene model track for nearby genes
gene_track <- GeneRegionTrack(
  esrra_neigh,
  showId = TRUE,
  name = "genes near ESRRA",
  geneSymbol = TRUE
)

plotTracks(gene_track)
```

Interpretation ?

## Plot window

```{r}
library(Gviz)

# Define the plotted genomic window
plot_region <- range(esrra_neigh)
plot_chr <- as.character(seqnames(plot_region))[1]
plot_from <- start(plot_region)
plot_to <- end(plot_region)

data.frame(chromosome = plot_chr, from = plot_from, to = plot_to)
```

The plotting window is the ESRRA neighborhood used earlier, so the visual view stays connected to the same annotation object

```{r}
# Add chromosome and coordinate tracks
idx_track <- IdeogramTrack(genome = "hg19", chromosome = plot_chr)
axis_track <- GenomeAxisTrack()

class(idx_track)
class(axis_track)
```

## Gviz with the annotated ERBS peaks

```{r}
# Use signal values from the annotated peaks when they are available
annotated_signal_track <- annotated_esrra_window_peaks
if ("signalValue" %in% names(mcols(annotated_signal_track))) {
  mcols(annotated_signal_track) <- S4Vectors::DataFrame(signalValue = mcols(annotated_signal_track)$signalValue)
  annotated_peak_track <- DataTrack(
    annotated_signal_track,
    data = "signalValue",
    type = c("h", "p"),
    name = "annotated ERBS signal",
    genome = "hg19",
    chromosome = plot_chr
  )
} else {
  annotated_peak_track <- AnnotationTrack(
    annotated_esrra_window_peaks,
    name = "annotated ERBS peaks",
    genome = "hg19",
    chromosome = plot_chr
  )
}

class(annotated_peak_track)
```

```{r}
# Add nearby gene models
gene_track <- GeneRegionTrack(
  esrra_neigh,
  showId = TRUE,
  name = "genes near ESRRA",
  geneSymbol = TRUE,
  genome = "hg19",
  chromosome = plot_chr
)

class(gene_track)
```

The gene track draws nearby transcript models, giving the ERBS signal a genomic annotation context.

```{r}
# Draw the annotated peaks with gene context
plotTracks(
  list(idx_track, axis_track, annotated_peak_track, gene_track),
  from = plot_from,
  to = plot_to,
  chromosome = plot_chr,
  background.title = "gray90",
  col.title = "black",
  cex.title = 0.8
)
```

This combined plot places the 6 annotated ERBS peaks above the ESRRA neighborhood gene models in a genome track view.

Interpretation ?

## Gviz with the official GM12878 object

```{r}
# Use signal values from GM12878 when they are available
gm12878_signal_track <- esrra_window_peaks
if ("signalValue" %in% names(mcols(gm12878_signal_track))) {
  mcols(gm12878_signal_track) <- S4Vectors::DataFrame(signalValue = mcols(gm12878_signal_track)$signalValue)
  gm12878_peak_track <- DataTrack(
    gm12878_signal_track,
    data = "signalValue",
    type = c("h", "p"),
    name = "GM12878 ERBS signal",
    genome = "hg19",
    chromosome = plot_chr
  )
} else {
  gm12878_peak_track <- AnnotationTrack(
    esrra_window_peaks,
    name = "GM12878 ERBS peaks",
    genome = "hg19",
    chromosome = plot_chr
  )
}

class(gm12878_peak_track)
```

```{r}
# Draw the official GM12878 peaks with gene context
plotTracks(
  list(idx_track, axis_track, gm12878_peak_track, gene_track),
  from = plot_from,
  to = plot_to,
  chromosome = plot_chr,
  background.title = "gray90",
  col.title = "black",
  cex.title = 0.8
)
```

Questions

-   Which peaks fall near ESRRA?

-   Are the strongest peaks inside a gene body, near a promoter, or distal?

-   How close are the peaks to neighboring genes?

-   What evidence would be needed to claim direct regulation?

-   Would this view change if the coordinates were interpreted as hg38 instead of hg19?
